Expanding Tidy Data Principles to Facilitate Missing Data Exploration, Visualization and Assessment of Imputations

### 2022-12-01 jss-copyedit/2022-01-12-4108/paper.tex
add files that have been copyedited by JSS

\documentclass[
]{jss}

@@

\author{
Nicholas Tierney\\Monash University \And Dianne Cook\\Monash University
}
\title{Expanding tidy data principles to facilitate missing data exploration, visualization and assessment of imputations}

\Plainauthor{Nicholas Tierney, Dianne Cook}
\Plaintitle{Expanding tidy data principles to facilitate missing data exploration, visualization and assessment of imputations}
\Shorttitle{Explore missingness with \pkg{naniar}}

\Abstract{
Despite the large body of research on missing value distributions and imputation, there is comparatively little literature with a focus on how to make it easy to handle, explore, and impute missing values in data. This paper addresses this gap. The new methodology builds upon tidy data principles, with the goal of integrating missing value handling as a key part of data analysis workflows. We define a new data structure, and a suite of new operations. Together, these provide a connected framework for handling, exploring, and imputing missing values. These methods are available in the \proglang{R} package \pkg{naniar}.
}

@@

\Address{
    }

@@

As data science has become a more solid field, theories and principles have developed to describe best practices. One such idea is ``tidy data'', which defines a clean, analysis-ready format that informs workflows converting raw data through a data analysis pipeline \citep{Wickham2014}. Another idea is the grammar of graphics, describing how to map data values into a visualization \citep{Wilkinson2012}. These principles have been widely adopted, but do not address the problem of missing data. In particular, there is little guidance on how to handle missing values in a data analysis workflow. Most analysis and visualization software simply drop missing values when making a plot, although some (\pkg{ggplot2}) provide a warning (Figure \ref{fig:warning}).

\begin{CodeChunk}
\begin{figure}

{\centering \includegraphics[width=0.75\linewidth]{tidy-missing-data-paper_files/figure-latex/warning-1}

}

\caption[How \pkg{ggplot2} behaves when displaying missing values]{How \pkg{ggplot2} behaves when displaying missing values. A warning message is displayed, but missing values are not shown in the plot.}\label{fig:warning}
\end{figure}
\end{CodeChunk}

@@

While there have been many software tools for exploring missing data, they do not work together. The graphics literature provides several solutions for exploring missings visually by incorporating them into the plot in some way. For example, imputing values to be 10\% below the minimum to include all observations into a scatter plot \citep{Cook2007}, or treating missing values as an equal category of data \citep{Unwin1996}. The ideas from the graphics literature need to be translated into tidy data tools to integrate missing data handling in a data analysis pipeline.

This paper is organized in the following way. Section \ref{background} provides the background to tidy data principles and tools (\ref{tidy-data-concepts}) and missing data representations (\ref{missing-data-rep-dep}). Section \ref{existing-software} summarises existing software for handling missing data. The new work extending the tidy tools to facilitate exploring, visualizing, and imputing missing data is discussed in Section \ref{extensions}. Graphics (\ref{graphics}) and Numerical summaries (\ref{num-sum}) of missing values are then discussed. An application illustrating the use of the new methods is shown in Section \ref{case-study}. Section \ref{discussion} discusses strengths, limitations, and future directions.

@@

Features of tidy data were formally described in \citet{Wickham2014}, and were discussed in terms of their importance for data science by \citet{Donoho2017}, and tools for data analysis. Tidy data is defined by \citet{r4ds} as:

\begin{quote}
\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  each variable must have its own column.
\item
  each observation must have its own row.
\item
  each value must have its own cell.
\end{enumerate}
\end{quote}

Tidy data is easier to work with and analyze because the variables are in the same format as they would be put into modelling software. This helps the analyst works swiftly and clearly, closing up opportunities for errors. Tidy data principles are general, but are comprehensively implemented in the \proglang{R} programming language, so this paper focuses on \proglang{R} \citep{rcore}.

@@

Data manipulation is made input- and output-tidy with \proglang{R} packages \pkg{dplyr} and \pkg{tidyr} \citep{dplyr, tidyr}. These provide the five ``verbs'' of data manipulation: data reshaping, sorting, filtering, transforming, and aggregating. Data reshaping goes from long to wide formats; sorting arranges rows in a specific order; filtering removes rows based on a condition; transforming, changes existing variables or adds new ones; aggregating creates a single value from many values, say, for example, in computing the minimum, maximum, and mean.

@@

Visualization tools only have tidy data as their input, as the output is a graphic. The popular domain specific language \pkg{ggplot2} maps variables in a dataset to features (referred to as aesthetics) of a graphic \citep{ggplot2}. For example, a scatterplot can be created by mapping two variables to the x and y axes, and specifying a point geometry.

@@

Modelling tools work well with tidy data, as they have a clear mapping from variables in the data to the formula for a model. For example in \proglang{R}, y regressed on x and z is: \code{lm(y ~ x + z)}. Modelling tools are input tidy, but their output is always messy - it is not in the right format for subsequent steps in analysis. For example, estimated coefficients, predictions, and residuals from one model cannot be easily combined with the output of another model. Messy models have been partially addressed with the \pkg{broom} package, which tidies up model outputs into a tidy data format for data analysis, and the developing \pkg{recipes} package, which helps make modelling input- and output-tidy \citep{recipes}.

@@

Current tools for missing data are messy. Missing data tools can be used to perform imputations, missing data diagnostics, and data visualizations. However, these tools suffer the same problems as modelling: they use tidy input, but produce messy output - their output is challenging to integrate with other steps of data analysis. The complex, often multivariate nature of imputation methods also makes them difficult to represent. Visualization methods for missing data do not map data features to the aesthetics of a graphic, as in \pkg{ggplot2}, limiting expressive exploration.

@@

The convention for representing missingness is a \textbf{b}inary matrix, \(B\), for data \(y\) with \(i\) rows and \(j\) columns:

\[
b_{ij} =\begin{cases}
1, & \text{if } y_{ij} \text{ is missing} \\
0, & \text{if } y_{ij} \text{ is observed}
\end{cases}
\]

There are many ways each value can be missing, we adopt the notation used in \citep{vanBuuren2018}. The information in \(B\) can be used to arrive at three categories of missing values: missing completely at random (MCAR), missing at random (MAR), and missing not at random (MNAR). The distribution of missing values in \(b_{ij}\) can depend on the entire dataset, represented as \(Y = (Y_{obs}, Y_{miss})\). This relationship can be defined by the \emph{missing data model} \(Pr(b_{ij} | Y_{obs}, Y_{miss}, \psi)\), the probability of missingness is conditional on data observed, data missing, and some probability parameter of missingness, \(\psi\). This helps to precisely define categories of missing values.

\textbf{MCAR} is where values being missing have no association with observed or unobserved data, that is, \(Pr(B = 1 | Y_{obs}, Y_{miss}) = Pr(B = 1 | \psi)\). Essentially, the probability of an observation being missing is unrelated to anything else, only the parameter \(\psi\), the overall probability of missingness. Although a convenient scenario, it is not actually possible to confirm, or clearly distinguish from MAR, as it relies on statements on data unobserved. In \textbf{MAR}, missingness only depends on data observed, not data missing, that is, \(Pr(B = 1 | Y_{obs}, Y_{miss}, \psi) = Pr(B = 1 | Y_{obs}, \psi)\). Some structure or dependence between missing and observed values is allowed, provided it can be explained by data observed, and some overall probability of missingness. In \textbf{MNAR}, missingness is related to values observed, and unobserved: \(Pr(B = 1 | Y_{obs}, Y_{miss}, \psi)\). This assumes conditioning on all observations: data goes missing due to some phenomena unobserved, including the structure of the missing data itself. This presents a challenge in analysis, as it is difficult to verify, and implies bias in analysis due to the unobserved phenomena. Visualizations can help assess whether data may be MCAR, MAR or MNAR.

@@

\pkg{VIM} \citep{VIM} implements well-used imputation methods: K nearest neighbors, regression, hot-deck, and iterative robust model-based imputation. These diverse approaches allow for imputing with semi-continuous, continuous, count, and categorical data. VIM identifies imputed cases by adding an indicator variable with a suffix ``\code{_imp}''. So, ``\code{Var1}'' has a sibling column, ``\code{Var1_imp}'', with values TRUE or FALSE indicating imputation. \pkg{VIM} also has a variety of visualization methods, discussed in Section \ref{exploration}. \pkg{simputation} provides an interface to imputation methods from \pkg{VIM}, in addition providing hotdeck imputation, and the EM algorithm \citep{simputation, Dempster1977}. \pkg{simputation} provides a consistent formula interface for all imputation methods, and always returns a dataframe with the updated imputations. \pkg{Hmisc} \citep{Hmisc}, provides predictive mean matching, \pkg{imputeTS} \citep{imputeTS}, provides time series imputation methods, and \pkg{missMDA} \citep{missMDA}, imputes data using principal component analysis.

@@

The \proglang{MANET} (Missings Are Now Equally Treated) software \citep{Unwin1996} focussed on multivariate categorical data, with missingness explicitly added as a category. \proglang{MANET} also provided univariate visualizations of missing data using linked brushing between a reference plot of the missingness for each variable, and a plot of the data as a histogram or barplot. The \proglang{MANET} software is no longer maintained and cannot be installed. The approach of \citep{Swayne1998} in the software \proglang{XGobi}, further developed in \proglang{ggobi} \citep{Cook2007}, focussed on multivariate quantitative data. Missingness is incorporated into plots in \proglang{ggobi} by setting them to be 10\% below the minimum value.

\pkg{MissingDataGUI} provides a Graphical User Interface (GUI) for exploring missing data structure, both numerically and visually. Using a GUI to explore missing data facilitates rapid insight into missingness structures. However, this comes as a trade off, as insights are not captured or recorded with a GUI, making it challenging to incorporate into reproducible analyses. This distracts and breaks analysis workflow, inviting mistakes.

\pkg{VIM} (Visualizing and Imputing Missing Data) provides visualization methods to identify and explore observed, imputed, and missing values. These include spinograms, spinoplots, missingness matrices, plotting missingness in the margins of other plots, and other summaries. However, these visualizations do not map variables to graphical aesthetics, creating friction when moving through analysis workflows, making them difficult to extend to new circumstances. Additionally, data used to create the visualizations cannot be accessed, posing a barrier to further exploration.

The \pkg{inspectdf} package provides tools for summarising, comparing, and visualising dataframes \citep{inspectdf}. In particular the \texttt{inspect\_na()} function provides a summary of missing values for variables. \texttt{inspect\_cat()} summarises the levels of categories in each column in a dataset. The \texttt{show\_plot()} function shows a higher lever overview of the contents of a dataframe and its categories, as well as missing values. All of these functions allow for comparing missing values across two dataframes. For example, comparing the missing values in two dataframes with the same column names.

\pkg{ggplot2} removes missing values with a warning (Figure \ref{fig:warning}), and only incorporates missingness into visualizations when mapping a discrete variable with missings to a graph aesthetic. This has some limitations, shown in Figure \ref{fig:gg-box-na}, a boxplot visualization of school grade and test scores. If there are missings in a continuous variable like test score, \pkg{ggplot2} omits the missings and prints a warning message. However, if a discrete variable like school year has missing values, an NA category is created for school year, where scores are placed.

\begin{CodeChunk}
\begin{figure}

{\centering \includegraphics[width=1\linewidth]{tidy-missing-data-paper_files/figure-latex/gg-box-na-1}

}

\caption[\pkg{ggplot2} provides different visualizations depending on what type of data has missing values for data of student test scores in school year]{\pkg{ggplot2} provides different visualizations depending on what type of data has missing values for data of student test scores in school year. (A) Data is complete; (B) Missings are only in year - an additional "NA" boxplot is created; (C) Missings only in scores, no additional missingness information is shown; (D) Missings in both scores and year, additional missing information is shown. The missingness category is only shown when there are missings in categorical variables such as year (plots (B) and (D)). In (C), no missingness information is given on the graphic, despite there being missings in score, and a warning message is displayed about the number of missing values omitted.}\label{fig:gg-box-na}
\end{figure}
\end{CodeChunk}

@@

Missing data can be explored and visualised in closed source software such as \proglang{STATA}, \proglang{SAS}, and \proglang{SPSS}. These provide basic table frequencies of missing data, and basic visualisations on missingness patterns. However, none of these software provide functions to specifically visualise missing values in bivariate or multivariate settings, and it is only possible to visualise missing values after performing data analysis, counting and summarising missingness. Imputation methods from mean to regression, EM, and multiple imputation are all available, with out of the box diagnostics provided to assist comparing imputations. \proglang{python} also provides exploratory data analysis for missing data within \pkg{missingno} \citep{Bilogur2018}, and the \pkg{scikit-learn} package provides imputation methods \citep{scikit-learn}.

@@

Applying \pkg{tidyverse} principles has the potential to clarify missing data exploration, visualization, and imputation. This section discusses how these principles are applied for data structures (\ref{data-structure}), common operations (verbs) (\ref{verbs}), graphics (\ref{graphics}), and data summaries (\ref{num-sum}). Care has been taken to make the names and design these features intuitive for their purpose, and is discussed throughout.

@@

We propose a new form for tidy representation of missing data based on these ideas from past research. Four features are added to the shadow matrix to facilitate analysis of missing data, illustrated in Figure \ref{fig:nabularfig}. These are briefly described below, and discussed in further detail in section \ref{nabular-data}.

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  \textbf{Missing value labels}. Simple labels for missing and not missing to clearly identify missing values for analysis and plotting. We propose ``NA'' and ``!NA''. (Figure \ref{fig:nabularfig}). This improves the 0 and 1 values in \(B\), which do not clearly identify missingness. These values follow a principle: ``\textbf{clarity of labelling}'' - the matrix's meaning is transparent, and anybody looking at these values could understand what they mean, which is not the case of binary values. Equally, these values could instead be ``missing'' or ``present''.
\item
  \textbf{Special missing values}: building on \textbf{missing value labels}, the values in the shadow matrix can be ``special'' missing values, indicated by ``NA\_\textless suffix\textgreater{}''. For example: ``NA\_instrument'' uses a short label, ``instrument'', indicating instrument error resulting in missing values. These could be also used to indicate imputations, (Figure \ref{fig:nabularfig}).
\item
  \textbf{Coordinated names}: variable names in the shadow matrix gain a consistent short suffix, "\_NA", keeping names coordinated throughout analysis (Figure \ref{fig:nabularfig}). It makes a clear distinction with ``var\_NA'' being a random variable of the missingness of a variable, ``var''. This suffix is short and easy to remember during data analysis, and shifts the focus from the value of a variable, to its missingness state. Although this might break naming conventions for lower case ``snake\_case'' variables (e.g., \citet{Wickham2014}), this distinguishes these columns from other columns.
\item
  \textbf{Connectedness}: binding the shadow matrix column-wise to the original data as a factor creates a single, connected, nabular data format, in sync with the data, at the cost of a larger file. It is useful for visualization, summaries, and tracking imputed values, discussed in more detail in \ref{nabular-data}.
\end{enumerate}

@@

\emph{Nabular} data binds the shadow matrix column-wise to the original data. It is a portmanteau of ``NA'' and ``tabular''. \emph{Nabular} data explicitly links missing values to data, keeping corresponding observations together and removing the possibility of mismatching records (Figure \ref{fig:nabularfig}). \emph{Nabular} data facilitates visualization and summaries by allowing the user to reference the missingness of a variable in a coordinated way: the missingness of ``var'', as ``var\_NA'' during analysis. \emph{Nabular} data is a snapshot of the missingness of the data. This means when \emph{nabular} data are imputed, those imputed values can easily be identified in analysis (``var'' vs ``var\_NA''). Imputing values on \emph{nabular} data automatically tracks these imputations. Together these features make it possible to do tasks that are not possible by simply adding a binary missing matrix, \(B\) during data analysis.

\begin{CodeChunk}
\begin{figure}

{\centering \includegraphics[width=1\linewidth]{/Users/njtierney/github/njtierney/papers/tidy-missing-data-paper/paper/images/diagram}

}

\caption[The process of creating nabular data]{The process of creating nabular data. Data transformed to shadow matrix, where values are either not missing or missing: "!NA" or "NA". The shadow matrix can be converted to long form to create missingness summary plots. Nabular data is created by binding the columns of the data and shadow matrix. Special missing values (such as -99) are identified as special missings, and values imputed and tracked. Nabular data can be used to identify imputations and explore data values alongside missings, providing a useful format for missing data exploration and analysis.}\label{fig:nabularfig}
\end{figure}
\end{CodeChunk}

\emph{Nabular} data is similar to classical data formats with quality or flag variables associated with each measured variable, e.g., Scripps CO2 data \citep{Keeling2005-scripps}, GHCN data \citep{Durre2008-ghcn}. Using additional columns to represent missingness information follows best practices for data organization, described in \citep{Ellis2018} and \citep{Broman2018}: (1) keep one thing in a cell and (2) describe additional features of variables in a second column. Here they suggest to indicate censored data with an extra variable called ``VariableNameCensored'', which would be TRUE if censored, otherwise FALSE. This information can now be represented in the shadow columns as special missing values. Encoding special missing values is achieved by defining logical conditions and suffixes. This is implemented with the \code{recode_shadow()} function in \pkg{naniar} (Discussed in Section \ref{verbs}).

>>>>

\documentclass[article]{jss}

\usepackage{thumbpdf,lmodern}
\graphicspath{{Figures/}} % folder with figures

@@

\author{
Nicholas Tierney\\Monash University, \\Telethon Kids Institute
\And Dianne Cook\\Monash University
}
\title{Expanding Tidy Data Principles to Facilitate Missing Data Exploration,
Visualization and Assessment of Imputations}

\Plainauthor{Nicholas Tierney, Dianne Cook}
\Plaintitle{Expanding Tidy Data Principles to Facilitate Missing Data Exploration,
Visualization and Assessment of Imputations}
\Shorttitle{Explore Missingness with \pkg{naniar}}

\Abstract{
Despite the large body of research on missing value distributions and
imputation, there is comparatively little literature with a focus on how to
make it easy to handle, explore, and impute missing values in data.  This
paper addresses this gap.  The new methodology builds upon tidy data
principles, with the goal of integrating missing value handling as a key
part of data analysis workflows.  We define a new data structure, and a
suite of new operations.  Together, these provide a connected framework for
handling, exploring, and imputing missing values.  These methods are
available in the \proglang{R}~package \pkg{naniar}. \citep{pkg:naniar}, available from the Comprehensive \proglang{R} Archive Network (CRAN) at \url{https://CRAN.R-project.org/package=naniar}.
}

@@

\Address{
  Nicholas Tierney\\
  Monash University\\
  Melbourne, Victoria, Australia\\
  \emph{and}\\
  Telethon Kids Institute\\
  Perth, Western Australia, Australia\\
  nicholas.tierney\@gmail.com

  Dianne Cook\\
  Monash University\\
  Melbourne, Victoria, Australia\\
  }

@@

As data science has become a more solid field, theories and principles have
developed to describe best practices.  One such idea is ``tidy data'', which
defines a clean, analysis-ready format that informs workflows converting raw
data through a data analysis pipeline \citep{Wickham2014}.  Another idea is
the grammar of graphics, describing how to map data values into a
visualization \citep{Wilkinson2012}.  These principles have been widely
adopted, but do not address the problem of missing data.  In particular,
there is little guidance on how to handle missing values in a data analysis
workflow.  Most analysis and visualization software simply drop missing
values when making a plot, although some (\pkg{ggplot2}) provide a warning
(Figure~\ref{fig:warning}).

\begin{figure}[t!]
\centering
\includegraphics[width=0.75\linewidth]{warning-1}
\caption[How \pkg{ggplot2} behaves when displaying missing values]{How
\pkg{ggplot2} behaves when displaying missing values.  A warning message is
displayed, but missing values are not shown in the plot.}\label{fig:warning}
\end{figure}

@@

While there have been many software tools for exploring missing data, they
do not work together.  The graphics literature provides several solutions
for exploring missings visually by incorporating them into the plot in some
way.  For example, imputing values to be 10\% below the minimum to include
all observations into a scatter plot \citep{Cook2007}, or treating missing
values as an equal category of data \citep{Unwin1996}.  The ideas from the
graphics literature need to be translated into tidy data tools to integrate
missing data handling in a data analysis pipeline. We have implemented these tools for exploring missing data into the \proglang{R} package \pkg{naniar} \citep{pkg:naniar}, available from the Comprehensive \proglang{R} Archive Network (CRAN) at \url{https://CRAN.R-project.org/package=naniar}.

This paper is organized in the following way.  Section~\ref{background}
provides the background to tidy data principles and tools
(Section~\ref{tidy-data-concepts}) and missing data representations
(Section~\ref{missing-data-rep-dep}).  Section~\ref{existing-software} summarizes
existing software for handling missing data.  The new work extending the
tidy tools to facilitate exploring, visualizing, and imputing missing data
is discussed in Section~\ref{extensions}.  Graphics (Section~\ref{graphics}) and
Numerical summaries (Section~\ref{num-sum}) of missing values are then discussed.
An application illustrating the use of the new methods is shown in
Section~\ref{case-study}.  Section~\ref{discussion} discusses strengths,
limitations, and future directions.

@@

Features of tidy data were formally described in \citet{Wickham2014}, and
were discussed in terms of their importance for data science by
\citet{Donoho2017}, and tools for data analysis.  Tidy data is defined by
\citet{r4ds} as:
%
% \begin{quote}
\begin{enumerate}
% \def\labelenumi{\arabic{enumi}.}
% \tightlist
\item
  each variable must have its own column.
\item
  each observation must have its own row.
\item
  each value must have its own cell.
\end{enumerate}
% \end{quote}
%
Tidy data is easier to work with and analyze because the variables are in
the same format as they would be put into modelling software.  This helps
the analyst works swiftly and clearly, closing up opportunities for errors.
Tidy data principles are general, but are comprehensively implemented in the
\proglang{R} programming language, so this paper focuses on \proglang{R}
\citep{rcore}.

@@

Data manipulation is made input- and output-tidy with \proglang{R}~packages
\pkg{dplyr} and \pkg{tidyr} \citep{dplyr, tidyr}.  These provide the five
``verbs'' of data manipulation: data reshaping, sorting, filtering,
transforming, and aggregating.  Data reshaping goes from long to wide
formats; sorting arranges rows in a specific order; filtering removes rows
based on a condition; transforming, changes existing variables or adds new
ones; aggregating creates a single value from many values, say, for example,
in computing the minimum, maximum, and mean.

@@

Visualization tools only have tidy data as their input, as the output is a
graphic.  The popular domain specific language \pkg{ggplot2} maps variables
in a dataset to features (referred to as aesthetics) of a graphic
\citep{ggplot2}.  For example, a scatterplot can be created by mapping two
variables to the $x$ and $y$~axes, and specifying a point geometry.

@@

Modelling tools work well with tidy data, as they have a clear mapping from
variables in the data to the formula for a model.  For example in
\proglang{R}, $y$ regressed on $x$ and $z$ is: \code{lm(y ~ x + z)}.  Modelling
tools are input tidy, but their output is always messy -- it is not in the
right format for subsequent steps in analysis.  For example, estimated
coefficients, predictions, and residuals from one model cannot be easily
combined with the output of another model.  Messy models have been partially
addressed with the \pkg{broom} package, which tidies up model outputs into a
tidy data format for data analysis, and the developing \pkg{recipes}
package, which helps make modelling input- and output-tidy \citep{recipes}.

@@

Current tools for missing data are messy.  Missing data tools can be used to
perform imputations, missing data diagnostics, and data visualizations.
However, these tools suffer the same problems as modelling: they use tidy
input, but produce messy output -- their output is challenging to integrate
with other steps of data analysis.  The complex, often multivariate nature
of imputation methods also makes them difficult to represent.  Visualization
methods for missing data do not map data features to the aesthetics of a
graphic, as in \pkg{ggplot2}, limiting expressive exploration.

@@

The convention for representing missingness is a \textbf{b}inary matrix,
\(B\), for data \(y\) with \(i\) rows and \(j\) columns:
%
\[
b_{ij} =\begin{cases}
1, & \text{if } y_{ij} \text{ is missing} \\
0, & \text{if } y_{ij} \text{ is observed}
\end{cases}
\]
%
There are many ways each value can be missing, we adopt the notation used in
\citep{vanBuuren2018}.  The information in \(B\) can be used to arrive at
three categories of missing values: missing completely at random (MCAR),
missing at random (MAR), and missing not at random (MNAR).  The distribution
of missing values in \(b_{ij}\) can depend on the entire dataset,
represented as \(Y = (Y_\text{obs}, Y_\text{miss})\).  This relationship can be
defined by the \emph{missing data model} \(\Prob(b_{ij} \mid Y_\text{obs}, Y_\text{miss},
\psi)\), the probability of missingness is conditional on data observed,
data missing, and some probability parameter of missingness, \(\psi\).  This
helps to precisely define categories of missing values.

MCAR is where values being missing have no association with
observed or unobserved data, that is, \(\Prob(B = 1 \mid Y_\text{obs}, Y_\text{miss}) = \Prob(B
= 1 \mid \psi)\).  Essentially, the probability of an observation being missing
is unrelated to anything else, only the parameter \(\psi\), the overall
probability of missingness.  Although a convenient scenario, it is not
actually possible to confirm, or clearly distinguish from MAR, as it relies
on statements on data unobserved.  In MAR, missingness only depends
on data observed, not data missing, that is, \(\Prob(B = 1 \mid Y_\text{obs}, Y_\text{miss},
\psi) = \Prob(B = 1 \mid Y_\text{obs}, \psi)\).  Some structure or dependence between
missing and observed values is allowed, provided it can be explained by data
observed, and some overall probability of missingness.  In MNAR,
missingness is related to values observed, and unobserved: \(\Prob(B = 1 \mid
Y_\text{obs}, Y_\text{miss}, \psi)\).  This assumes conditioning on all observations:
data goes missing due to some phenomena unobserved, including the structure
of the missing data itself.  This presents a challenge in analysis, as it is
difficult to verify, and implies bias in analysis due to the unobserved
phenomena.  Visualizations can help assess whether data may be MCAR, MAR or
MNAR.

@@

\pkg{VIM} \citep{VIM} implements well-used imputation methods: $K$ nearest
neighbors, regression, hot-deck, and iterative robust model-based
imputation.  These diverse approaches allow for imputing with
semi-continuous, continuous, count, and categorical data.  \pkg{VIM} identifies
imputed cases by adding an indicator variable with a suffix \code{_imp}.
So, \code{Var1} has a sibling column, \code{Var1_imp}, with values
\code{TRUE} or \code{FALSE} indicating imputation.  \pkg{VIM} also has a variety of
visualization methods, discussed in Section~\ref{exploration}.
\pkg{simputation} provides an interface to imputation methods from
\pkg{VIM}, in addition providing hotdeck imputation, and the
Expectation-maximization (EM) algorithm
\citep{simputation, Dempster1977}.  \pkg{simputation} provides a consistent
formula interface for all imputation methods, and always returns a dataframe
with the updated imputations.  \pkg{Hmisc} \citep{Hmisc}, provides
predictive mean matching, \pkg{imputeTS} \citep{imputeTS}, provides time
series imputation methods, and \pkg{missMDA} \citep{missMDA}, imputes data
using principal component analysis.

@@

The \proglang{MANET} (missings are now equally treated) software
\citep{Unwin1996} focused on multivariate categorical data, with
missingness explicitly added as a category.  \proglang{MANET} also provided
univariate visualizations of missing data using linked brushing between a
reference plot of the missingness for each variable, and a plot of the data
as a histogram or barplot.  The \proglang{MANET} software is no longer
maintained and cannot be installed.  The approach of \citep{Swayne1998} in
the software \proglang{XGobi}, further developed in \proglang{ggobi}
\citep{Cook2007}, focused on multivariate quantitative data.  Missingness
is incorporated into plots in \proglang{ggobi} by setting them to be 10\%
below the minimum value.

\pkg{MissingDataGUI} \citep{pkg:MissingDataGUI}
provides a Graphical User Interface (GUI) for exploring
missing data structure, both numerically and visually.  Using a GUI to
explore missing data facilitates rapid insight into missingness structures.
However, this comes as a trade off, as insights are not captured or recorded
with a GUI, making it challenging to incorporate into reproducible analyses.
This distracts and breaks analysis workflow, inviting mistakes.

\pkg{VIM} (visualizing and imputing missing data) provides visualization
methods to identify and explore observed, imputed, and missing values.
These include spinograms, spinoplots, missingness matrices, plotting
missingness in the margins of other plots, and other summaries.  However,
these visualizations do not map variables to graphical aesthetics, creating
friction when moving through analysis workflows, making them difficult to
extend to new circumstances.  Additionally, data used to create the
visualizations cannot be accessed, posing a barrier to further exploration.

The \pkg{inspectdf} package provides tools for summarizing, comparing, and
visualizing dataframes \citep{inspectdf}.  In particular the
\code{inspect\_na()} function provides a summary of missing values for
variables.  \code{inspect\_cat()} summarizes the levels of categories in
each column in a dataset.  The \code{show\_plot()} function shows a higher
lever overview of the contents of a dataframe and its categories, as well as
missing values.  All of these functions allow for comparing missing values
across two dataframes.  For example, comparing the missing values in two
dataframes with the same column names.

\pkg{ggplot2} removes missing values with a warning
(Figure~\ref{fig:warning}), and only incorporates missingness into
visualizations when mapping a discrete variable with missings to a graph
aesthetic.  This has some limitations, shown in Figure~\ref{fig:gg-box-na},
a boxplot visualization of school grade and test scores.  If there are
missings in a continuous variable like test score, \pkg{ggplot2} omits the
missings and prints a warning message.  However, if a discrete variable like
school year has missing values, an NA category is created for school year,
where scores are placed.

\begin{figure}[t!]
\centering
\includegraphics[width=1\linewidth]{gg-box-na-1}
%
\caption[\pkg{ggplot2} provides different visualizations depending on what
type of data has missing values for data of student test scores in school
year]{\pkg{ggplot2} provides different visualizations depending on what type
of data has missing values for data of student test scores in school year.
(A) Data is complete; (B) Missings are only in year -- an additional ``NA''
boxplot is created; (C) Missings only in scores, no additional missingness
information is shown; (D) Missings in both scores and year, additional
missing information is shown.  The missingness category is only shown when
there are missings in categorical variables such as year (plots (B) and
(D)).  In (C), no missingness information is given on the graphic, despite
there being missings in score, and a warning message is displayed about the
number of missing values omitted.}\label{fig:gg-box-na}
%
\end{figure}

@@

Missing data can be explored and visualized in closed source software such
as \proglang{STATA}, \proglang{SAS}, and \proglang{SPSS}.  These provide
basic table frequencies of missing data, and basic visualizations on
missingness patterns.  However, none of these software provide functions to
specifically visualize missing values in bivariate or multivariate settings,
and it is only possible to visualize missing values after performing data
analysis, counting and summarizing missingness.  Imputation methods from
mean to regression, EM, and multiple imputation are all available, with out
of the box diagnostics provided to assist comparing imputations.
\proglang{Python} also provides exploratory data analysis for missing data
within \pkg{missingno} \citep{Bilogur2018}, and the \pkg{scikit-learn}
package provides imputation methods \citep{scikit-learn}.

@@

Applying \pkg{tidyverse} principles has the potential to clarify missing
data exploration, visualization, and imputation.  This section discusses how
these principles are applied for data structures (Section~\ref{data-structure}),
common operations (verbs) (Section~\ref{verbs}), graphics (Section~\ref{graphics}), and data
summaries (Section~\ref{num-sum}).  Care has been taken to make the names and design
these features intuitive for their purpose, and is discussed throughout.

@@

We propose a new form for tidy representation of missing data based on these
ideas from past research.  Four features are added to the shadow matrix to
facilitate analysis of missing data, illustrated in
Figure~\ref{fig:nabularfig}.  These are briefly described below, and
discussed in further detail in Section~\ref{nabular-data}.
%
\begin{enumerate}
% \def\labelenumi{\arabic{enumi}.}
%
\item
  \textbf{Missing value labels}:  Simple labels for missing and not missing
to clearly identify missing values for analysis and plotting.  We propose
``NA'' and ``!NA'' (Figure~\ref{fig:nabularfig}).  This improves the 0 and
1 values in \(B\), which do not clearly identify missingness.  These values
follow a principle: ``clarity of labelling'' -- the matrix's meaning
is transparent, and anybody looking at these values could understand what
they mean, which is not the case of binary values.  Equally, these values
could instead be ``missing'' or ``present''.
%
\item
  \textbf{Special missing values}: Building on \textbf{missing value
labels}, the values in the shadow matrix can be ``special'' missing values,
indicated by ``NA\_\textless suffix\textgreater{}''.  For example:
``NA\_instrument'' uses a short label, ``instrument'', indicating instrument
error resulting in missing values.  These could be also used to indicate
imputations (Figure~\ref{fig:nabularfig}).
%
\item
  \textbf{Coordinated names}: Variable names in the shadow matrix gain a
consistent short suffix, ``\_NA'', keeping names coordinated throughout
analysis (Figure~\ref{fig:nabularfig}).  It makes a clear distinction with
``var\_NA'' being a random variable of the missingness of a variable,
``var''.  This suffix is short and easy to remember during data analysis,
and shifts the focus from the value of a variable, to its missingness state.
Although this might break naming conventions for lower case ``snake\_case''
variables \citep[e.g.,][]{Wickham2014}, this distinguishes these columns from
other columns.
%
\item
  \textbf{Connectedness}: Binding the shadow matrix column-wise to the
original data as a factor creates a single, connected, nabular data format,
in sync with the data, at the cost of a larger file.  It is useful for
visualization, summaries, and tracking imputed values, discussed in more
detail in Section~\ref{nabular-data}.
\end{enumerate}

@@

\emph{Nabular} data binds the shadow matrix column-wise to the original
data.  It is a portmanteau of ``NA'' and ``tabular''.  \emph{Nabular} data
explicitly links missing values to data, keeping corresponding observations
together and removing the possibility of mismatching records
(Figure~\ref{fig:nabularfig}).  \emph{Nabular} data facilitates
visualization and summaries by allowing the user to reference the
missingness of a variable in a coordinated way: the missingness of ``var'',
as ``var\_NA'' during analysis.  \emph{Nabular} data is a snapshot of the
missingness of the data.  This means when \emph{nabular} data are imputed,
those imputed values can easily be identified in analysis (``var'' vs
``var\_NA'').  Imputing values on \emph{nabular} data automatically tracks
these imputations.  Together these features make it possible to do tasks
that are not possible by simply adding a binary missing matrix, \(B\) during
data analysis.

\begin{figure}[t!]
\centering
\includegraphics[width=1\linewidth]{diagram}
\caption[The process of creating nabular data]{The process of creating
nabular data.  Data transformed to shadow matrix, where values are either
not missing or missing: ``!NA'' or ``NA''.  The shadow matrix can be converted
to long form to create missingness summary plots.  Nabular data is created
by binding the columns of the data and shadow matrix.  Special missing
values (such as ``$-99$'') are identified as special missings, and values imputed
and tracked.  Nabular data can be used to identify imputations and explore
data values alongside missings, providing a useful format for missing data
exploration and analysis.}\label{fig:nabularfig}
\end{figure}

\emph{Nabular} data is similar to classical data formats with quality or
flag variables associated with each measured variable, e.g.,~Scripps CO2
data \citep{Keeling2005-scripps}, GHCN data \citep{Durre2008-ghcn}.  Using
additional columns to represent missingness information follows best
practices for data organization, described in \citep{Ellis2018} and
\citep{Broman2018}: (1) keep one thing in a cell and (2) describe additional
features of variables in a second column.  Here they suggest to indicate
censored data with an extra variable called ``VariableNameCensored'', which
would be TRUE if censored, otherwise FALSE.  This information can now be
represented in the shadow columns as special missing values.  Encoding
special missing values is achieved by defining logical conditions and
suffixes.  This is implemented with the \code{recode_shadow()} function in
\pkg{naniar} (discussed in Section~\ref{verbs}).

Special missing values are not a new idea, and have been implemented in
other statistical programming languages, \proglang{SPSS}, \proglang{SAS},
and \proglang{STATA}.  These typically represent missing values as a
full-stop, ``.'', and record special missing values as ``.a'' -- ``.z''.
These special values from these languages break the tidy principle of one
column having one type of value, as they record both the value, and the
multivariate missingness state.
